---
title: "covariate extraction"
output: html_document
date: "2026-02-22"
---

**Purpose:** Pull down covariate data & wrangle into panel model data structure

**Tasks**
- import data, wrangle into monthly values, extract bird info, and format bird data into year-month-cluster df

**Files**
- data/processed/bird_local/bird_panel_fullCA.csv = year-month-cluster for ALL CA counties (-99 for LA/Orange)
- 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## upload necessary packages
library(tidyverse)
library(sf)
library(ggplot2)
library(googledrive)
library(exactextractr)

## upload station data
cluster_shp <- st_read("data/processed/trapstations/cluster_all_centroidsbuffer/cluster_all_centroidsbuffer.shp")
cluster_info <- read.csv("data/processed/trapstations/stations_all_info.csv")

## upload ca shapefile
options(tigris_use_cache = TRUE)
ca <- tigris::counties("ca", class = "sf")
```


# Bird community competence
https://drive.google.com/drive/u/0/folders/1FgDkejWqE12rlvV5OiGxAyov5qkYbhq2

### A. Import data
```{r}
# authenticate google drive folder 
# drive_auth()

# direct to specific folder
folder <- drive_get("bird community")

# make loop to call in multiple csvs
read_drive_tif <- function(file_name, file_id) {
  temp_path <- tempfile(fileext = ".tif") # not the pngs
  drive_download(as_id(file_id), path = temp_path, overwrite = TRUE)
  r <- terra::rast(temp_path)
  message(paste("loaded:", file_name))
  return(r)
}

# show me which files i've called in (this is a bit overkill for 3 files)
g <- drive_ls(folder) %>% 
  filter(grepl("\\.tif", name, ignore.case = TRUE))

files <- setNames(g$id, g$name)

# make loop to give the object names
for(name in names(files)) {
  obj_name <- gsub("\\.tif$", "", name)
  assign(obj_name, read_drive_tif(name, files[[name]]), envir = .GlobalEnv)
}

```

### B. Wrangle data
```{r}
## add all rasters to 1 list

# a. create named list of all years
year_rasters <- lapply(names(files), function(name) {
  obj_name <- gsub("\\.tif$", "", name)
  get(obj_name, envir = .GlobalEnv)
})

names(year_rasters) <- gsub("\\.tif$", "", names(files))

# b. convert weekly (52 layers) rasters to monthly (12 layers) values
# assign 4–5 weeks per month
weeks_to_month <- list(month_01 = 1:4,
                       month_02 = 5:8,
                       month_03 = 9:13,
                       month_04 = 14:17,
                       month_05 = 18:21,
                       month_06 = 22:26,
                       month_07 = 27:30,
                       month_08 = 31:35,
                       month_09 = 36:39,
                       month_10 = 40:44,
                       month_11 = 45:48,
                       month_12 = 49:52)


# c. aggregate to months
monthly_rasters <- lapply(year_rasters, function(r) {
  
  # for each month, computer mean across week
  monthly <- lapply(weeks_to_month, function(weeks){
    terra::app(r[[weeks]], fun = mean, na.rm = TRUE)})
  
  # combine into 1 spatraster per year
  terra::rast(monthly)})

```

### C. Merge bird spatial data with cluster polygons
```{r}
## a. deal with CRS
cluster_vect <- vect(st_transform(cluster_shp, crs(monthly_rasters[[1]])))

## b. write funciton to extract motnly averages per cluster
extract_monthly_bird <- function(monthly_rasters, cluster_vect) {
  
  # store
  results <- list()
  
  # for each year in the raster stack
  for(year_name in names(monthly_rasters)) {
    
    r <- monthly_rasters[[year_name]] #rembered went from 52 to 12 layer raster
    
    # extract bird community competence mean
    avg_vals <- terra::extract(r, cluster_vect, fun = mean, na.rm = TRUE)
    
    # replace NA (polygons outside raster with -99) this is temporary until i figure out which mos to keep
    # mostly LA, Orange, Riverside counties
    avg_vals[is.na(avg_vals)] <- -99
    
    # story with cluster_id
    avg_vals_df <- data.frame(cluster_id = cluster_shp$cluster_id,
                              year = year_name,
                              avg_vals)
    
    results[[year_name]] <- avg_vals_df
  }
  
  # combine all years
  do.call(rbind, results)
}


## c. run the loop
bird_cluster <- extract_monthly_bird(monthly_rasters, cluster_vect)
```

### D. Clean up data for panel models
```{r}
## convert from wide to long to match panel data
bird_panel <- data.frame(bird_cluster, row.names = NULL) %>% 
  mutate(year = as.numeric(substr(as.character(year), nchar(year)-3, nchar(year)))) %>% 
  pivot_longer(cols = starts_with("month_"),
               names_to = "month",
               values_to = "bird_competence") %>% 
  mutate(month = sub("month_", "", month)) %>% 
  dplyr::select(cluster_id, year, month, bird_competence)
  
# save raw data before any spatial filtering is made
#write.csv(bird_panel, "data/processed/bird_local/bird_panel_fullCA.csv", row.names = FALSE)


## check to see how it merges with cluster info
bird_panel %>% 
  group_by(cluster_id) %>% 
  slice(1) %>% # only take 1 obs since bird is for all year-months
  left_join(cluster_info, by = "cluster_id") %>% 
  # check if -99 is applied to the right places outsdie CV
  filter(bird_competence == -99 & 
           county != "Los Angeles" & county != "Riverside" & 
           county != "Orange" & county != "Placer" &
           county != "Shasta")

## okay so it looks like the culprit counties make up majority of -99 (null values bc bird rasters dont exist there)
# theres a couple sac/yuba but i'm not too worried 

```


# Temperature 
```{r}



```

